Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-06-11 paper/paper.tex
fixed broken figure layout
typpass 2.0 over FEP

\begin{wrapfigure}{r}{.3\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action by following some policy, receives a consequential reward and observes the next state. The process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}
The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensors.
\citep{starkweather2017dopamine}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

@@

\section{Relation to existing cognitive accounts on default mode function}
\subsection{Predictive coding hypothesis}
Predictive coding mechanisms
\citep{clark2013whatever, friston2008hierarchical}
are a frequently evoked idea in the context of default mode function
\citep{bar2007}.
Cortical responses are explained as
emerging from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
Feed-forward sensory processing is constantly calibrated by
top-down modulation from more multi-sensory and associative brain regions
further away from primary sensory cortical regions.
The dynamic interplay between cortical processing levels
may enable inferences about the world by reconciling
gaps between fresh sensory input and expectations computed
based on stored prior information.
At each stage of neural processing,
an internally generated prediction of aspects of environmental sensations is
directly compared against the actual environmental input.
A prediction error at one of the processing levels
incurs plasticity changes of neuronal projections
(i.e., adapting model parameters)
to allow for gradually improved future prediction of the environment.
In this way,
the predictive coding hypothesis offers explanations for
the constructive, non-deterministic nature of sensory perception
\citep{friston2010free, buzsaki2006rhythms} and
the intimate relation of motor action to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
Contextual integration of sensorimotor perception-action cycles
may be maintained by top-down modulation
using a-priori information about the environment.

@@

\subsection{The free-energy principle and active inference}
% INTRO
According to the free-energy principle (FEP) and
active inference theories \citep{friston2010free, fristonAIorRL},
the brain corresponds to
a biomechanical inference engine. It is
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --or more precisely, an upper bound thereof-- \textit{relative} to the expectations about the external world derived from the brain's internal representations. The brain
would continuously generate hypothetical explanations of the world
and predict its sensory input.
Precursors of this theory can be traced back to Dayan and colleagues \citep{dayan1995helmholtz}
who introduced the \textit{Helmholtz machine}, a hierarchical factorial directional deep belief network. According to the FEP account,
the goal of the brain is to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ of objects in the world, their interactions and dynamics. The model then allows
to minimize surprise when these representations are confronted with sensory input $\x$ during action-perception cycles (i.e., the \textit{generative} model).
The FEP also postulates a companion model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$ to accomplish approximate inference. Put differently, the recognition model dreams imaginary worlds $\z$,
while the generative model tries to produce sensations $\x$ which match these
imagined mental scenarios.
Howewer, surprise is challenging to optimize numerically
because we need to sum over all hidden causes of the sensations.
Instead, FEP therefore minimizes an upper-bound of surprise,
namely the \text{free-energy} $F_G^R(\x)$ given by

@@

% CRITICISM
Despite its popularity, criticism against the FEP has arisen over the years,
some of which we outline in the following.
The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As previously noted \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it postulates a pair
of forward (generation) and backward (recognition) models that
together do not correspond to optimization of (a bound of) the marginal likelihood.
Thus, the brain may therefore be unlikely to implement a variant of the wake-sleep algorithm.
We note that the recent theory of
\textit{variational auto-encoders} (VAEs) \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm. This is because VAEs
overcome a number of the technical limits of the wake-sleep algorithm
by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free-energy, VAEs can be efficiently trained via back-propagation of prediction errors.

Since FEP-based theories \citep{friston2010free,fristonAIorRL} conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
in this environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
\begin{wrapfigure}{r}{.45\linewidth}
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a closed space is used as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}
\end{wrapfigure}
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddles to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would readily
identify the visual-input-free room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply finding the room exit.

Finally, we note that FEP and active inference can be recast in our model-free RL framework.
This becomes possible by postulating that the reward signal maximized by the DMN is
the negative surprise,
that is, the log-likehood of sensory the current priors the agent has about the world. More explicitly, this corresponds to

\begin{equation}
  r^{\text{FEP}} = \underbrace{\log (p_G)}_{-\text{FEP generative surprisal}}.
    \end{equation}
% \paragraph{Comparison to our proposed theory.}
% On the surface, a common point between the FEP and our proposed RL-based framework
%   is placing the minimization of a surprise signal at the core of brain function.
%   Indeed in RL, surprise minimization is subsumed  by accurate prediction of
%   rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
%   \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}.
Such a surprisal-based reinforcement learning scheme has previously been advocated \citep{schmidhuber2010,mohamed2015} under the equivalent framework of information compression.
Nevertheless, minimization of surprise alone
may be insufficient to explain behavior observed in humans and other intelligent animals.
% The dark-room experiment, for example, provides a formidable argument against in favor of the insufficiency of the FEP and active inference theories to explain intelligent behavior in humans: when a human is put in a dark room, they don't just sit down in a corner forever, what would be the FEP-optimal thing to do; they actively search for the light switch, and start exploring the depths of the room, possibly looking for food or the way out!

>>>>

The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensors.
\citep{starkweather2017dopamine}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

\begin{wrapfigure}{r}{.3\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action by following some policy, receives a consequential reward and observes the next state. The process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}

@@

\section{Relation to existing functional accounts}
\subsection{Predictive coding hypothesis}
Predictive coding mechanisms
\citep{clark2013whatever, friston2008hierarchical}
are a frequently evoked idea in the context of default mode function
\citep{bar2007}.
Cortical responses are explained as
emerging from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
Feed-forward sensory processing is constantly calibrated by
top-down modulation from more multi-sensory and associative brain regions
further away from primary sensory cortical regions.
The dynamic interplay between cortical processing levels
may enable inferences about the world by reconciling
gaps between fresh sensory input and expectations computed
based on stored prior information.
At each stage of neural processing,
an internally generated prediction of aspects of environmental sensations is
directly compared against the actual environmental input.
A prediction error at one of the processing levels
incurs plasticity changes of neuronal projections
(i.e., adapting model parameters)
to allow for gradually improved future prediction of the environment.
In this way,
the predictive coding hypothesis offers explanations for
the constructive, non-deterministic nature of sensory perception
\citep{friston2010free, buzsaki2006rhythms} and
the intimate relation of motor action to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
Contextual integration of sensorimotor perception-action cycles
may be maintained by top-down modulation
using a-priori information about the environment.

@@

\subsection{The free-energy principle and active inference}
% INTRO
According to theories of the free-energy principle (FEP) and
active inference \citep{friston2010free, fristonAIorRL},
the brain corresponds to
a biomechanical reasoning engine. It is
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --more precisely, an upper bound thereof-- \textit{relative} to the expectations about the external world derived from internal representations. The brain
would continuously generate hypothetical explanations of the world
and predict its sensory input.
Precursors of this interpretation can be traced back to Dayan and colleagues \citep{dayan1995helmholtz}
who introduced the \textit{Helmholtz machine}, a hierarchical factorial directional deep belief network. According to the FEP account,
the goal of the brain is to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ of objects in the world, their interactions and dynamics. The model then allows
to minimize surprise when these representations are confronted with sensory input $\x$ during action-perception cycles (i.e., the \textit{generative} model).
The FEP also proposes a companion model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$ to accomplish approximate inference. Put differently, the recognition model dreams imaginary worlds $\z$,
while the generative model tries to produce sensory sensations $\x$ which match these
imagined mental scenarios.
Howewer, surprise is challenging to optimize numerically
because we need to sum over all hidden causes of the sensations.
Instead, FEP therefore minimizes an upper-bound of surprise,
namely the \text{free energy} $F_G^R(\x)$ given by

@@

% CRITICISM
Despite its popularity, criticism against the FEP has arisen over the years,
some of which outlined in the following.
The main algorithm for minimizing free energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As these authors noted,
a crucial drawback of the wake-sleep algorithm is that it involves a pair
of forward (generation) and backward (recognition) models that
together do not correspond to optimization of (a bound of) the marginal likelihood.
The brain may therefore be unlikely to implement a variant of the wake-sleep algorithm.
The recent theory of
\textit{variational auto-encoders} (VAEs) \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm. VAEs
overcome a number of the technical limits of the wake-sleep algorithm
by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free energy, VAEs can be efficiently trained via back-propagation of prediction errors.

Since theories based on the FEP \citep{friston2010free,fristonAIorRL}
conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
given the environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
\begin{wrapfigure}{r}{.45\linewidth}
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a light-deprived closed space can be used
  as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}
\end{wrapfigure}
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddle points to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would readily
identify the visual-input-deprived room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply finding the room exit.

Finally, we note that FEP and active inference can be reframed in terms of
our model-free RL framework.
This becomes possible by recasting the reward signal maximized by the DMN to 
correspond to negative surprise,
that is, the log-likehood of current sensory priors
the agent has about the world. More explicitly, this corresponds to

\begin{equation}
  r^{\text{FEP}} = \underbrace{\log (p_G)}_{-\text{FEP generative surprise}}.
    \end{equation}
% \paragraph{Comparison to our proposed theory.}
% On the surface, a common point between the FEP and our proposed RL-based framework
%   is placing the minimization of a surprise signal at the core of brain function.
%   Indeed in RL, surprise minimization is subsumed  by accurate prediction of
%   rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
%   \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}.
Such a surprise-guided reinforcement learning scheme has previously been advocated
under the equivalent framework of information compression \citep{schmidhuber2010,mohamed2015}.
Nevertheless, minimization of surprise quantities alone
may be insufficient to explain
the diversity of behaviors observed in humans and other intelligent animals.
% The dark-room experiment, for example, provides a formidable argument against in favor of the insufficiency of the FEP and active inference theories to explain intelligent behavior in humans: when a human is put in a dark room, they don't just sit down in a corner forever, what would be the FEP-optimal thing to do; they actively search for the light switch, and start exploring the depths of the room, possibly looking for food or the way out!
